# Poverty measurement methodology 
***

The World Bank set a target of reducing extreme poverty to [less than 3 percent by 2030](https://sdg-tracker.org/no-poverty) and to ensure continued focus and steady progress toward the goal, the institution set an interim target of 9 percent by 2020. Progress towards this goal is measured by monitoring the share of the global population living below the international poverty line, currently set at US$1.90 in 2011 purchasing power parity (PPP) dollars. All eight countries in South Asia measure the international extreme poverty status of an individual by comparing consumption expenditures per capita against this poverty line. 

In this book, we  overview the data availability of household consumption surveys in the region of South Asia and the procedure followed for poverty measurement. The starting point in monitoring progress in poverty reduction and enhancing shared prosperity in the region is to have household consumption survey data that are not only available at reasonably frequent intervals, but are also comparable over time (@serajuddin2015data). Several spatial and temporal adjustments are necessary to take into account differences in costs of living. 

First, a consumer price index (CPI) is necessary to measure by how much the general price level in a country has changed over time. Deflating all nominal expenditures to real expenditures allows to ensure that welfare comparisons between two periods are not being driven by inflation. Figure \@ref(fig:cpi) displays how the CPI has changed for each country in South Asia over time. 

```{r cpi, echo=FALSE, warning=FALSE, eval=TRUE, error=FALSE, results='markup', tidy=TRUE, cache=FALSE,  fig.cap="[Consumer Price Index (CPI)](https://tab.worldbank.org/#/site/WBG/views/SAR_MNA_Summary/LineChart)"}
library(readr)
cpi <- read_csv("./tables/cpi.csv",col_names = TRUE, col_types = cols())

library(ggplot2)
ggplot(data=cpi, aes(x=year, y=cpi, group=country)) +
  geom_line(aes(linetype=country))+
  geom_point(aes(shape=country))
```

Second, spatial price differences, typically between urban and rural areas, can also be large and it is important to take them into account. In South Asia, price levels for rural households are often lower than for urban households. The ideal way to control for spatial differences in prices is to use a Paasche or Laspeyres index to account for differences in the cost of living across space. 

Third, it is necessary to convert expenditures to a common currency taking into account purchasing power. At present, data on purchasing power parity (PPP) comes from the [International Comparison Program (ICP)](http://www.worldbank.org/en/programs/icp) collected in 2011, which is absent for Afghanistan. Purchasing power parities (PPPs) are the rates of currency conversion that equalize the purchasing power of different currencies by eliminating the differences in price levels between countries. The PPP between countries A and B measures the number of units of country A’s currency required to purchase a basket of goods or services in country A as compared to one unit of country B’s currency to purchase a similar basket of goods in country B. This indicator is measured in terms of national currency per US dollar and shown in the following table:

|   Country   	|      Currency      	| ICP 2011 	|
|:-----------:	|:------------------:	|:--------:	|
| Afghanistan 	|   Afghan afghani   	|     NA  	|
|  Bangladesh 	|  Bangladeshi taka  	|  24.8493 	|
|    Bhutan   	| Bhutanese ngultrum 	| 16.96292 	|
|    India    	|    Indian rupee    	| 13.98707 	|
|   Maldives  	|  Maldivian rufiyaa 	| 10.67606 	|
|    Nepal    	|   Nepalese rupee   	| 25.75928 	|
|   Pakistan  	|   Pakistani rupee  	| 25.41426 	|
|  Sri Lanka  	|  Sri Lankan rupee  	| 42.21894 	|

A fourth adjustment is dividing total household expenditure by some measure of the number of people in the household, and to assign the resulting per capita welfare measure to each household member as an individual. Later in the applications, we show how larger households typically have lower per capita expenditure levels than smaller households. In South Asia, the consequences of dividing total household expenditures by a greater number of individuals would not be complete without considering the extent of economies of scale within the household and a discussion of how much children and the elderly typically consume compared to adults. 

The FGT measures devised by @fgt remain the most commonly used to measure poverty. A headcount ratio may be calculated in STATA as in the example below:

```markdown
# Example: Calculating poverty headcounts with 1.90 poverty line

# Calculate per capita expenditures by dividing total expenditures 
# by the number of household members 
  gen welfare=total_exp/hsize 

# Calculate real per capita expenditures by turning nominal  
# per capita expenditures into 2011 purchasing power parity (PPP) dollars
  welfare_ppp = welfare/cpi2011/icp2011/365*12

# Generate a dummy variable to determine whether real per capita expenditures
# fall below the 1.90 2011 PPP dollars poverty line
  gen poor190 = welfare_ppp < 1.90

# Calculate the weighted average of the dummy variable to measure the proportion
# of the population that falls below the 1.90 2011 PPP dollars poverty line
  sum poor190 [aw=wgt]
```
```{r,eval=FALSE,echo=FALSE}
The FGT measures devised by Foster, Greer and Thorbecke (1984)  remain the most commonly used to measure poverty. They are defined as:

$$\begin{equation}
P_{\alpha}=\frac{1}{N}\sum^{N}_{i=1}I_{i}\left(\frac{\bar{u}-u_{i}}{\bar{u}}\right)^{\alpha}
\end{equation}$$

where N is the sample size, $$\bar{u}$$ is the scalar-valued poverty line, $$u_{i}$$ is the flow-based measure of welfare (income, expenditures, assets), $$I_{i}$$ is an indicator variable taking value one if $u_{i} < \bar{u}$ and zero otherwise, and $$\alpha$$ is a parameter reflecting the weight placed on the severity of poverty. Setting $$\alpha =0$$ yields the poverty headcount ratio $P_{0}$ (the share of a population falling below the poverty line). The higher order measures, $$P_{1}$$ and $$P_{2}$$, yield the poverty gap measure (the money metric measure of the average financial transfer needed to bring all poor households up to the poverty line) and the squared poverty gap (an indicator of the severity of poverty that is sensitive to the distribution of well-being among the poor). As we progress from $$P_{0}$$ to $$P_{1}$$ to $$P_{2}$$, the $$P_{\alpha}$$ measure gets more and more sensitive to extremely low incomes.
```

The variable `welfare` contains the monthly nominal expenditures per capita in local currency. Once turned into daily 2011 USD (PPP), these values may be compared over time and between countries as in the following dashboard. The area to the left of the US$1.90 extreme poverty line represents the proportion of individuals living in extreme poverty. 

<script type='text/javascript' src='https://tab.worldbank.org/javascripts/api/viz_v1.js'></script><div class='tableauPlaceholder' style='width: 700px; height: 827px;'><object class='tableauViz' width='700' height='827' style='display:none;'><param name='host_url' value='https%3A%2F%2Ftab.worldbank.org%2F' /> <param name='embed_code_version' value='3' /> <param name='site_root' value='&#47;t&#47;WBG' /><param name='name' value='SAR_MNA_Summary&#47;Distribution' /><param name='tabs' value='no' /><param name='toolbar' value='yes' /><param name='showAppBanner' value='false' /><param name='filter' value='iframeSizedToWindow=true' /></object></div>
